---
title: "The Gender Pay Gap in UK Universities"
author: "Group 1 - Ankit Ray, Nitin Mathew, Sharan Pullabhatla, Vyshnav Varma"
date: "31/Oct/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE)
```

```{r Library Load}
library(tidyverse)
library(rmarkdown)
```

```{r Data Import, cache = TRUE}
# Import data
gpd_2122 <- read_csv("../data/UK Gender Pay Gap Data - 2021 to 2022.csv")
gpd_2021 <- read_csv("../data/UK Gender Pay Gap Data - 2020 to 2021.csv")
gpd_1920 <- read_csv("../data/UK Gender Pay Gap Data - 2019 to 2020.csv")
gpd_1819 <- read_csv("../data/UK Gender Pay Gap Data - 2018 to 2019.csv")
gpd_1718 <- read_csv("../data/UK Gender Pay Gap Data - 2017 to 2018.csv")
uni <- read_csv("../data/universities.csv")

# -------*--------
# Combine data
# Combine gpd
gpd <- rbind(gpd_2122, gpd_2021, gpd_1920, gpd_1819, gpd_1718)
# Join the universities data. Use inner join to drop any rows not in Uni
# Remove the EmployerName from GPD data as it is available in uni data
payGapData <- gpd %>%
  select(-EmployerName) %>%
  inner_join(uni, by =c("EmployerId"))
# Create a new column for assessment year by extracting DueDate's year
payGapData <- payGapData %>%
  mutate(year = case_when(str_sub(DueDate, 1, 4) == '2022' ~ '2021/22',
                          str_sub(DueDate, 1, 4) == '2021' ~ '2020/21',
                          str_sub(DueDate, 1, 4) == '2020' ~ '2019/20',
                          str_sub(DueDate, 1, 4) == '2019' ~ '2018/19',
                          str_sub(DueDate, 1, 4) == '2018' ~ '2017/18'))

```
---

# Changes in Gender Pay Gap for the period 2017-22
> While we see improvements over a 5 year period, there continues to be 
disparity in pay, more so for universities established prior to 1992

```{r Plot 1, fig.align='center', fig.width=10, fig.cap = "Figure 1", cache=TRUE}
# Figure 1
payGapData %>%
  # Limits columns to only required, will use data for all 5 years
  select(EmployerName, DiffMeanHourlyPercent, pre92, year) %>%
  mutate(uni_type = recode_factor(pre92, 
                                 yes = "Pre '92",
                                 no = "Post '92")) %>%
  ggplot(aes(x = year, y = DiffMeanHourlyPercent / 100, fill = uni_type)) +
  # Remove outliers as they are not relevant for this plot
  geom_boxplot(outlier.shape=NA) +
  # For the appearance of the plot background
  theme_bw() +
  # Customise the scaling displayed for y-axis
  scale_y_continuous(labels = scales::percent) + 
  # Labels
  labs(x = "", y = "Avg difference in hourly pay", 
       fill ="University\nCategory",
       title = "Gender Pay Gap in UK universities split by Pre and Post '92 universities",
       caption = "Source: UK Government Gender Pay Data, 2018 to 2022") +
  # Annotation for UoS
  # Refer https://r-graph-gallery.com/42-colors-names.html for colour names
  annotate("point", x = "2017/18", y = 19 / 100,
           colour = "cornflowerblue", size = 2) +
  annotate("point", x = "2018/19", y = 17.9 / 100,
           colour = "cornflowerblue", size = 2) +
  annotate("point", x = "2019/20", y = 17.8 / 100,
           colour = "cornflowerblue", size = 2) +
  annotate("point", x = "2020/21", y = 16.5 / 100,
           colour = "cornflowerblue", size = 2) +
  annotate("point", x = "2021/22", y = 15.8 / 100,
           colour = "cornflowerblue", size = 2) +
  # For an arrow pointing to the last UoS annotation
  annotate(
    geom = "curve", x = 5.2, y = 22 / 100, xend = "2021/22", yend = 16 / 100, 
    colour = "cornflowerblue", curvature = .1, 
    arrow = arrow(length = unit(2, "mm"))
  ) +
  # For text for UoS
  annotate(geom = "text", x = 5, y = 24 / 100, label = "University of\n   Sheffield", 
           colour = "cornflowerblue", hjust = "left", size = 3) +
  #For pre92 colour
  # Refer https://r-graphics.org/recipe-colors-palette-discrete for palletes
  scale_fill_brewer(palette = "Paired", breaks=c("Pre '92", "Post '92")) 
```

The data for Figure 1 is as per the **DiffMeanHourlyPercent** column from the
published reports. This column represents the **Mean % difference between male 
and female hourly pay (negative = women's mean hourly pay is higher)**. The 
outliers have been excluded from this plot.

Figure 1 shows how the official gender pay gap has changed from 2017/18 up 
until 2021/22. Looking at the interquartile ranges, we see them approximately 
between **9% and 23%** for the first period and improving slightly to 
around **8% and 21%** for the last fully published period.

Additionally, we do see a noticeable difference in the gap when looking at the Pre 
and Post '92 universities with the **Post '92** universities performing better
over these years.

Finally, looking at the **University of Sheffield**, we find it having slight
improvements in line with other Pre '92 universities while remaining at the lower 
end of the 1st quartile for this category of universities for the reported 
periods.

### Number of submissions at a glance
```{r No. of Submissions}
(gpd_cont <- payGapData %>%
   mutate(`University Category` = recode_factor(pre92, 
                                                yes = "Pre '92",
                                                no = "Post '92")) %>%
   mutate(`Year of Submission` = year)) %>%
  group_by(`Year of Submission`, `University Category`) %>%
  summarise("No. of Submissions" = n()) %>%
  pivot_wider(names_from = `University Category`,
              values_from = `No. of Submissions`) %>%
  knitr::kable(caption="Table 1 - Shows how many universities have submitted GPD data", 
               align='c')
```
To provide additional context to the graph in Figure 1, we have Table 1 above 
showing the number of universities which have submitted their GPD data for the 5 
year period in scope of this report.

---

# Data Sources
* The primary data for the report is based on the UK Government's official GPD 
data available from https://gender-pay-gap.service.gov.uk/viewing/download
* The universities have been identified using the data available at https://vle.shef.ac.uk/bbcswebdav/pid-6298247-dt-content-rid-45156957_1/xid-45156957_1